Too hard to get: the role of probabilistic expectations and cognitive complexity in destructive multi-dimensional reference points

### 2025-02-27 manuscript-apa.qmd
Aspen Han Xiang Yu - Project submission #1 for grading - @nrdowling @lim1an

My research is theoretically founded on the @koszegi.2006.modelreferencedependenta model of reference-dependence (henceforth KR model). Reference points have redefined preference modeling in economics. Introduced as a core component of Kahneman and Tversky's [-@kahneman.1979.prospecttheory, -@tversky.1991.lossaversion] prospect theory, it posits that people evaluate outcomes relative to a reference point rather than on absolute terms and weight losses more than gains. However, they did not identify the source of reference points, which became a source of contention. The KR model endogenises the reference points to be the agent’s (rational) expectations, specifically his/her probabilistic beliefs held in the recent past about what will or should happen. This accommodated alternative arguments about the origins of reference points, such as the status quo [e.g. @genesove.2001.lossaversion; @kahneman.1990.experimentaltests] and refutes to it [e.g. @plott.2005.willingnesspay; @tversky.1991.lossaversion]. Pinpointing the source of reference points was a major contribution as it allowed for more detailed studies into their effects and design, which motivates my use of the KR model as a theoretical baseline. The KR model also partially reconciles the EU theorem with prospect theory as KR considers the utility of a realized outcome to be the sum of both neoclassical consumption utility (absolute outcome levels) and gain-loss utilities (relative outcome levels), and it weights outcomes by their objective probabilities. This enables the KR model to satisfy internal consistency axioms such as transitivity which strengthens its normative appeal. However, the KR model, similar to most if not all reference point and EU models, also assumes that utilities across different dimensions of consumption are additively separate, which I seek to challenge. Yet, it seems unrealistic to think that people would view reference points in isolation from one another and determine how much to work towards each with complete disregard for the others.

@@

Subjects were randomly assigned to four treatment groups, which varied in terms of the presence of reference points (i.e. targets) in speed and accuracy, the probabilities of their work being assessed by strict or lenient assessment criterion, which respectively caught all or caught only half of mistakes made, and the extent of explanation of the reference points. Treatment 1 was the control with no reference points (and hence no explanation) and certainty of being assessed by a strict criterion. Treatments 2, 3, and 4 were set the same reference points and primed. Treatment 2 had a 75% probability of getting a lenient assessment criteria and 25% probability of strict, whereas treatment 3 had the inverse. Additionally, these two treatments will explain to subjects how they should work to achieve both targets concurrently. Treatment 4 will have the same criteria assignment probabilities as treatment 2 but without the explanation of reference points. Subjects were primed to think that it would be manageable to concurrently achieve both targets under the lenient criterion but difficult to do so under the strict criterion.

@@

```{r}
#| label: read-in-data

@@

@@

$p(e)$ is the level payoff from effort exertion, summed across both dimensions. $c(e)$ is the cost of effort. $μ_1[(n(e_1 )-N)\mathbb{I}(n<N) + λ_1(n(e_1)-N)\mathbb{I}(n≤N)]$ is the gain-loss utility in the speed dimension, where $μ_1≥0$ is the gain-loss parameter, $λ≥1$ is the loss aversion parameter, and $\mathbb{I}(.)$ is an indicator function equaling 1 when the condition in the bracket holds and 0 otherwise. $μ_2 [(Q-q(e_2))\mathbb{I}(q<Q) + λ_2(Q-q(e_2))\mathbb{I}(q≥Q) ] )]$ is analogously defined for the accuracy dimension.

@@

\begin{align}
U = &p(e_1, e_2) - c(e_1, e_2) + \nonumber \\
    &E[\mathbb{P}(\{n \geq N-\varepsilon \} \cap \{q \leq Q+ε\}] \times \theta \times \nonumber \\
    &\{\mu_1[(n(e_1 )-N)\mathbb{I}(n>N) + \lambda_1(n(e_1)-N)\mathbb{I}(n \leq N)] + \nonumber \\
    &\mu_2[(Q-q(e_2))\mathbb{I}(q<Q) + \lambda_2(Q-q(e_2))\mathbb{I}(q \geq Q)]\} \nonumber
\end{align}

The first additional term $E[\mathbb{P}(\{n \geq N-\varepsilon \} \cap \{q \leq Q+ε\}]$ captures the agent’s expected probability of simultaneously achieving (within some bandwidth $\varepsilon$ of) all reference points. When this expected probability is lower, the agent weights the gain-loss utilities less and hence is less responsive to the reference points. The second additional parameter $\theta≥0$ is a parameter decreasing in the cognitive complexity required to integrate the multiple reference points, so greater cognitive complexity attenuates reference point effects.

@@

>>>>

My research is theoretically founded on the @koszegi.2006.modelreferencedependenta model of reference-dependence (henceforth KR model). Reference points have redefined preference modeling in economics. Introduced as a core component of Kahneman and Tversky's [-@kahneman.1979.prospecttheory; -@tversky.1991.lossaversion] prospect theory, it posits that people evaluate outcomes relative to a reference point rather than on absolute terms and weight losses more than gains. However, they did not identify the source of reference points, which became a source of contention. The KR model endogenises the reference points to be the agent’s (rational) expectations, specifically his/her probabilistic beliefs held in the recent past about what will or should happen. This accommodated alternative arguments about the origins of reference points, such as the status quo [e.g. @genesove.2001.lossaversion; @kahneman.1990.experimentaltests] and refutes to it [e.g. @plott.2005.willingnesspay; @tversky.1991.lossaversion]. Pinpointing the source of reference points was a major contribution as it allowed for more detailed studies into their effects and design, which motivates my use of the KR model as a theoretical baseline. The KR model also partially reconciles the EU theorem with prospect theory as KR considers the utility of a realized outcome to be the sum of both neoclassical consumption utility (absolute outcome levels) and gain-loss utilities (relative outcome levels), and it weights outcomes by their objective probabilities. This enables the KR model to satisfy internal consistency axioms such as transitivity which strengthens its normative appeal. However, the KR model, similar to most if not all reference point and EU models, also assumes that utilities across different dimensions of consumption are additively separate, which I seek to challenge. Yet, it seems unrealistic to think that people would view reference points in isolation from one another and determine how much to work towards each with complete disregard for the others.

@@

Subjects were randomly assigned to four treatment groups, which varied in terms of the presence of reference points (i.e. targets) in speed and accuracy, the probabilities of their work being assessed by strict or lenient assessment criterion, which respectively caught all or caught only half of mistakes made, and the extent of explanation of the reference points. Treatment 1 was the control with no reference points (and hence no explanation) and certainty of being assessed by a strict criterion. Treatments 2, 3, and 4 were set the same reference points and primed. Treatment 2 had a 75% probability of getting a lenient assessment criteria and 25% probability of strict, whereas treatment 3 had the inverse. Additionally, these two treatments explained to subjects how they should work to achieve both targets concurrently. Treatment 4 will have the same criteria assignment probabilities as treatment 2 but without the explanation of reference points. Subjects were primed to think that it would be manageable to concurrently achieve both targets under the lenient criterion but difficult to do so under the strict criterion.

@@

```{r}
#| label: read-in-main-data
#| include: false

source("data\\main-data-cleaning.R")

@@

There are `r nrow(slider_data_main)` number of observations.

@@

$p(e)$ is the level payoff from effort exertion, summed across both dimensions. $c(e)$ is the cost of effort. $\mu_1[(n(e_1 )-N)\mathbb{I}(n>N) + \lambda_1(n(e_1)-N)\mathbb{I}(n \leq N)]$ is the gain-loss utility in the speed dimension, where $\mu_1≥0$ is the gain-loss parameter, $\lambda≥1$ is the loss aversion parameter, and $\mathbb{I}(.)$ is an indicator function equaling 1 when the condition in the bracket holds and 0 otherwise. $\mu_2 [(Q-q(e_2))\mathbb{I}(q<Q) + \lambda_2(Q-q(e_2))\mathbb{I}(q \geq Q) ] )]$ is analogously defined for the accuracy dimension.

@@

\begin{align}
U = &p(e_1, e_2) - c(e_1, e_2) + \nonumber \\
    &E[\mathbb{P}(\{n \geq N-\varepsilon \} \cap \{q \leq Q+\varepsilon\}] \times \theta \times \nonumber \\
    &\{\mu_1[(n(e_1 )-N)\mathbb{I}(n>N) + \lambda_1(n(e_1)-N)\mathbb{I}(n \leq N)] + \nonumber \\
    &\mu_2[(Q-q(e_2))\mathbb{I}(q<Q) + \lambda_2(Q-q(e_2))\mathbb{I}(q \geq Q)]\} \nonumber
\end{align}

The first additional term $E[\mathbb{P}(\{n \geq N-\varepsilon \} \cap \{q \leq Q+\varepsilon\}]$ captures the agent’s expected probability of simultaneously achieving (within some bandwidth $\varepsilon$ of) all reference points. When this expected probability is lower, the agent weights the gain-loss utilities less and hence is less responsive to the reference points. The second additional parameter $\theta≥0$ is a parameter decreasing in the cognitive complexity required to integrate the multiple reference points, so greater cognitive complexity attenuates reference point effects.

@@

## C. Reduced form and structural estimation of loss aversion
